Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-09-21 analysis/paper/paper.qmd
finished editing the text

Therefore, it can be inferred that the Qianshangying hominins deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes. The technological strategy reflected in the Qianshangying faceted spheroids includes the use of a conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy. This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes rather than the shaping of a cobble into a predetermined morphology [@inizan1999; @duke2021].

@@

These findings also contribute to the long-standing debate over the nature of spheroids-whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies. In the case of the Qianshangying faceted spheroids, the evidence strongly supports the latter. Faceted spheroids recovered from other Early Pleistocene sites in Barranco Leon and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020]. These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology. It is important to note, however, that many spheroids from East Africa may not fall into this category. As noted by @mora2005, these quartz spheroids may have been produced through percussive techniques. They are often covered with battering marks, exhibit smoother surfaces and lack ridges, features that suggest they may be better classified as bolas within a typological framework [@leakey1971]. It may be more appropriate to describe these artifacts in technological terms, rather than relying on morphology-based typologies that can lead to confusion. For instance, referring to them as battering spheroids could help distinguish them from faceted spheroids.

Another issue we aimed to address is the relationship between polyhedrons, spheroids, and bolas. While no bolas were identified at Qianshangying, our quantitative analyses clearly differentiated polyhedrons from spheroids or subspheroids. These two artifact types differ in size, morphology, and production sequence. Polyhedrons at Qianshangying represent heavily reduced cores, whose edge angle increase results from extensive flake removal. Their primary purpose lies in producing flakes, placing them within a flaking strategy. In contrast, spheroids and subspheroids were intentionally shaped as end products, reflecting a shaping strategy. These two technological approaches of flaking and shaping strategies coexisted at Qianshangying, indicating a diversity of reduction goals.

@@

## QSY spheroids in the context of East Asia

@@

Over 1522 spheroids have been identified across more than 23 archaeological sites or localities in East Asia However, many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited. Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each. Qianshangying is typical with 'r n_spheroids' spheroids. By contrast, three sites-Xujiayao (XJY), Dingcun (DC), and JST—have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun]. Notably, XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].

@@

We found six sites exhibiting similar strategies to QSY. For example, at DC and MJQ, thick and flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun]. MJQ specimens retain parts of the cortex, while at DC, flakes with large exterior and small interior platform angles were removed from blanks to increase edge angles and enhance roundness [@Tao1984Dingcun; @Wei1984]. Among the six sites, all but MJQ are located within the Fen River and Lishui River regions. These sites were grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by trihedral picks [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe]. The trihedral picks were produced by striking the dorsal surfaces of large flakes to concentrate edges at the tip, resulting in triangular outlines and robust pointed ends, classifying them as large cutting tools (LCTs)[@kuman2014large; @li2018currently; @yang2016late]. This is a significant observation, as it implies that spheroid production techniques similar to those at Qianshangying are more frequently associated with lithic assemblages characterized by intentional shaping strategies.

It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Acheulean technology. We infer that faceted spheroids in China appear to represent a shaping tradition that shares the same crucial technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], and this explains why they often occur in Acheulean techno-complex. We now find that faceted spheroids occur at QSY within the small flake tool industry of northern China. Since this industry was previously regarded as a simple core and flake technology, the presence of faceted spheroids suggests that it is more complex and exhibits greater variability than previously thought.

@@

Quantitative approaches to studying spheroids remain limited [@Muller2023; @deweyer2017; @tittonSubspheroidsLithicAssemblage2020]. We propose a complete, efficient, reproducible and transparent methodological framework here based on spherical harmonics to quantitatively assess morphological variation among flaked pieces, allowing for the distinction between cores and spheroids and evaluating the degree of morphological standardization in spheroids. Past approaches to analyzing the morphology of flaked pieces have been limited, with geometric morphometrics being the last major method innovation [@hallinan2025; @cao2025; @lycett2013]. However, this method has significant constraints, including the requirement for homologous landmarks, which limits its applicability to artifacts like cores of various shapes and spheroids. Spherical harmonics offer a viable alternative, showing strong potential for precision and interpretability. We can reasonably expect broader applications of this method across a range of archaeological artifacts in the future.

@@

This paper also presents one of the first systematic overviews of the spatial and temporal distribution and technological variability of spheroids within China. We found that Chinese spheroids display considerable diversity in raw materials, production techniques, and morphology. Some assemblages contain spheroids that may be comparable to those from Qianshangying and tend to appear within Acheulean or shaping-oriented techno-complexes, often dated later than Qianshangying. This highlights the particular significance of the Qianshangying spheroids. However, the current dataset remains limited, and many sites lack detailed morphological and technological studies of spheroids, hampering broader inferences about hominin technological behaviors. We therefore emphasize the need for site-specific analysis of spheroids in future research.

>>>>

Therefore, our results indicate that the Qianshangying hominins deliberately employed a standardized production approach to create spheroids with consistent and formalized shapes. The technological strategy reflected in the Qianshangying faceted spheroids includes the use of a conceptual template, standardized production, and the achievement of a target product, which together demonstrate that this was a shaping strategy. This stands in contrast to flaking strategies, where the primary objective is the removal of usable flakes rather than the shaping of a cobble into a predetermined morphology [@inizan1999; @duke2021].

@@

These findings also contribute to the long-standing debate over the nature of spheroids--whether they should be interpreted as byproducts of flake production, or as targeted end products of shaping strategies. In the case of the Qianshangying faceted spheroids, the evidence strongly supports the latter. Faceted spheroids recovered from other Early Pleistocene sites in Barranco Leon and ‘Ubeidiya have also been reported to show evidence consistent with shaping strategy [@Muller2023; @tittonSubspheroidsLithicAssemblage2020]. These intentionally shaped spheroids appear contemporaneous with, or even predate, the emergence of Acheulean technology. It is important to note, however, that many spheroids from East Africa may not fall into this category. As noted by @mora2005, these quartz spheroids may have been produced through percussive techniques. They are often covered with battering marks, exhibit smoother surfaces and lack ridges, features that suggest they may be better classified as bolas within a typological framework [@leakey1971]. It may be more appropriate to describe these artifacts in technological terms, rather than relying on morphology-based typologies that can lead to confusion. For instance, referring to them as battering spheroids could help distinguish them from faceted spheroids.

Another question our results are relevant to is the relationship between polyhedrons, spheroids, and bolas. While no bolas were identified at Qianshangying, our quantitative analyses clearly differentiated polyhedrons from spheroids or subspheroids. These two artifact types differ in size, morphology, and production sequence. Polyhedrons at Qianshangying represent heavily reduced cores, whose edge angle increase results from extensive flake removal. Their primary purpose lies in producing flakes, placing them within a flaking strategy. In contrast, spheroids and subspheroids were intentionally shaped as end products, reflecting a shaping strategy. These two technological approaches of flaking and shaping strategies coexisted at Qianshangying, indicating a diversity of reduction goals.

@@

## Qianshangying spheroids in the context of East Asia

@@

Over 1522 spheroids have been identified across more than 23 archaeological sites or localities in East Asia However, many of these findings derive from surface surveys rather than systematic excavations, and the recovered quantities are often limited. Thus, the number of spheroids per site varies considerably: 19 sites have yielded fewer than ten specimens each. Qianshangying is typical with six spheroids and seven subspheroids. By contrast, three sites--Xujiayao (XJY), Dingcun (DC), and JST—-have produced substantial assemblages, each exceeding 100 spheroids [@Wang2016Houjiayao; @Wang2010Jinstai; @SXA2014Dingcun]. Notably, XJY alone has yielded over 1,000 specimens [@Wang2016Houjiayao].

@@

We found six sites exhibiting similar strategies to Qianshangying. For example, at DC and MJQ, thick and flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun]. MJQ specimens retain parts of the cortex, while at DC, flakes with large exterior and small interior platform angles were removed from blanks to increase edge angles and enhance roundness [@Tao1984Dingcun; @Wei1984]. Among the six sites, all but MJQ are located within the Fen River and Lishui River regions. These sites were grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by trihedral picks [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe]. The trihedral picks were produced by striking the dorsal surfaces of large flakes to concentrate edges at the tip, resulting in triangular outlines and robust pointed ends, classifying them as large cutting tools (LCTs)[@kuman2014large; @li2018currently; @yang2016late]. This is a significant observation, as it implies that spheroid production techniques similar to those at Qianshangying are more frequently associated with lithic assemblages characterized by intentional shaping strategies.

It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Acheulean technology. We infer that faceted spheroids in China appear to represent a shaping tradition that shares the same crucial technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], and this explains why they often occur in Acheulean techno-complex. We now find that faceted spheroids occur at Qianshangying within the small flake tool industry of northern China. Since this industry was previously regarded as a simple core and flake technology, the presence of faceted spheroids suggests that it is more complex and exhibits greater variability than previously thought.

@@

Quantitative approaches to studying spheroids remain limited [@Muller2023; @deweyer2017; @tittonSubspheroidsLithicAssemblage2020]. We propose a complete, efficient, reproducible and transparent methodological framework here based on spherical harmonics to quantitatively assess morphological variation among flaked pieces, allowing for the distinction between cores and spheroids and evaluating the degree of morphological standardization in spheroids. Past approaches to analyzing the morphology of flaked pieces have been limited, with geometric morphometrics being the most recent major method innovation [@hallinan2025; @cao2025; @lycett2013]. However, this method has significant constraints, including the requirement for homologous landmarks, which limits its applicability to artifacts like cores of various shapes and spheroids. Spherical harmonics offer a viable alternative, showing strong potential for precision and interpretability. We can reasonably expect broader applications of this method across a wide range of archaeological artifacts in the future.

@@

This paper also presents one of the first systematic overviews of the spatial and temporal distribution and technological variability of spheroids within East Asia. We found that East Asian spheroids display considerable diversity in raw materials, production techniques, and morphology. Some assemblages contain spheroids that may be comparable to those from Qianshangying and tend to appear within Acheulean or shaping-oriented techno-complexes, often dated later than Qianshangying. This highlights the particular significance of the Qianshangying spheroids. However, the current dataset remains limited, and many sites lack detailed morphological and technological studies of spheroids, hampering broader inferences about hominin technological behaviors. We therefore emphasize the need for site-specific analysis of spheroids in future research.
